---
title: "Publishing Reports Using R and R Markdown"
author: "Kristina Riemer"
date: "10/27/2021"
output: html_document
---

## Overview

- End goal: each participant have a published Rmarkdown report on their research
- Steps: 
  - Intros
  - Defining Rmarkdown and markdown
  - Creating example Rmarkdown
  - Turning R script into Rmarkdown
  - Options for publishing
  - Publish created Rmarkdown

## Curriculum & notes

#### 1. Rmarkdown introduction


start with showing what all can be done using Rmarkdowns as the basis
  - example Rmd with chunks of code and text and outputs (tables, figures)
  - document and share code and outputs, for your future self and others


1. Code chunks of other programming languages like Python
2. Create books with R package `bookdown`
3. Create websites with R package `blogdown`
4. Make presentations
5. Automatically generate in-text citations from bibliography
6. Include interactive components with R package `htmlwidgets` or with Shiny
7. Create dashboard using R package `flexdashboard` (run through example at end)

Start with walking through how to create a basic one



- Walkthrough of [Data Carpentry for Biologists knitr lecture](https://datacarpentry.org/semester-biology/materials/knitr/)
  - Can also add images, math equations, and tables to text using markdown

Final content from example Rmarkdown: 

````
---
title: "Rodent Report"
author: "Kristina Riemer"
date: "10/27/2021"
output: html_document
---

## Purpose

Exploring rodent population data at the **Portal Project**. 

List of tasks: 

1. Load data from [database](http://figshare.com/articles/Portal_Project_Teaching_Database/1314459)
2. Clean data into time series
3. Make plot

## Read in packages

```{r, message=FALSE}`r ''`
library(dplyr)
library(ggplot2)
```

## Read in data

```{r}`r ''`
data <- read.csv("https://ndownloader.figshare.com/files/2292172")
head(data)
```

The data includes `r knitr::inline_expr('length(unique(data$species_id))')` species. 

## Clean up data and plot

Summarizing data for number of occurrences per year and per species. Removed zero values and NAs, and species x because they went extinct. 

```{r}`r ''`
#generate time series data with dplyr
time_series <- data %>% 
  group_by(species_id, year) %>% 
  summarize(count = n())
head(time_series)
```

```{r, message=FALSE, echo=FALSE}`r ''`
ggplot(time_series, aes(x = year, y = count)) +
  geom_line() +
  facet_wrap(~species_id)
```

````

#### 2. Publish Rmarkdown

- Publish so that have a publicly available document with a URL that can be shared with anyone
- Two methods with their own pros and cons
  - RStudio Connect
    - Need license, which UA got this year
    - Limited to research-related products (no class assignments or toy examples)
    - Can add collaborators to update
    - Can schedule regular updates and send automated emails about them to list of email addresses
  - RPubs
    - Free for anyone to use
    - Can post any kind of content, not limited to research
    - Can't collaboratively update
    - Updating can be a bit finicky
- Walkthrough of publishing through RStudio Connect
  - Register UA account for RStudio Connect
  - Connect RStudio to RStudio Connect
    - Tools -> Global Options -> Publishing -> Connect...
    - Input UA server address -> Next
    - Some folks had to install R packages before RStudio Connect was an option
  - Click blue publish button while .Rmd is open and select RStudio Connect option
  - Select "Publish document with source code" (takes longer if big files, but renders on the fly) or "Publish finished document only" (faster but requires most recently knit html)
  - Show [example published Rmarkdown](https://viz.datascience.arizona.edu/connect/#/apps/61e325a4-b005-4cc6-b090-fbb2d939429e/access/3)
- Walkthrough of publishing through RPubs


#### 3. Create new Rmarkdown

- Purpose is to create an Rmarkdown from existing R code or script
- Spend half an hour working, discussing, and troubleshooting
- Folks discussed what they were going to work on



#### 4. Example of a use for Rmarkdown beyond simple report (optional)



Example of dashboard with Shiny component using rodent example data:

````
---
title: "Rodent Dashboard"
author: "Kristina Riemer"
date: "10/27/2021"
output: flexdashboard::flex_dashboard
runtime: shiny
---

```{r}`r ''`
library(flexdashboard)
library(dplyr)
library(ggplot2)
library(shiny)
```

```{r}`r ''`
data <- read.csv("https://ndownloader.figshare.com/files/2292172")
time_series <- data %>% 
  group_by(species_id, year) %>% 
  summarize(count = n())
```

Column 2
--------------------------------------------------

### Number of species

```{r}`r ''`
valueBox(length(unique(data$species_id)), icon = "fa-tag")
```

### Time series plot

```{r}`r ''`
ggplot(time_series, aes(x = year, y = count)) +
  geom_line() +
  facet_wrap(~species_id)
```

Column 1
--------------------------------------------------

### About rodents

This is some example content from the **Portal Project** to demonstrate how to create a dashboard from an Rmarkdown. 

### Rodent dataset by year

```{r}`r ''`
numericInput("chosen_year", "Input year", 1980)

renderTable({
  head(filter(data, year == input$chosen_year))
})
```

````



## Resources

- [Workshop recording](https://youtube.com/watch?v=WaX2uocJ_2M)
- [R Markdown book](https://bookdown.org/yihui/rmarkdown/)
- [Markdown syntax](https://www.markdownguide.org/basic-syntax/)
- [Markdown cheat sheet](https://www.markdownguide.org/cheat-sheet/)
- [GitHub-flavored markdown](https://guides.github.com/features/mastering-markdown/)
- [Rmarkdown cheat sheet](https://raw.githubusercontent.com/rstudio/cheatsheets/master/rmarkdown.pdf)
- [UA RStudio Connect signup](https://datascience.arizona.edu/RSConnect)
- [RStudio Connect user guide](https://docs.rstudio.com/connect/1.7.4/user/index.html)
- [RStudio Connect server](https://viz.datascience.arizona.edu/)

## Help

Feel free to drop by the [CCT Data Science Team office hours](https://datascience.cals.arizona.edu/office-hours), which happen every Tuesday morning. We would love to help you with your Rmarkdown-specific, and general R, questions! 

You can also [make an appointment with Kristina](https://calendar.google.com/calendar/u/0/selfsched?sstoken=UU83LWZEMmhrdWxUfGRlZmF1bHR8N2NkYWViOThjOGZlZjFhMTQyNzIyNDIxYjJkODc0Yjg) to discuss this content and get troubleshooting help. 
