---
title: "Weather Data in R"
output: github_document
urlcolor: blue
---

## Workshop intro

Date: 6/30/2021
Time: 11am - 1pm (AZ)

We will record this workshop and make the recording available on YouTube. 

### Who are we

We are in the Data Science Institute at UA. Our group is called DIAG, we work on computational tools and training for agricultural and ecological data. We want to enable researchers with their work, through training and consulting, so let us know if we could help you in any way! 

Kristina and David introduce ourselves. 

### Workshop overview/logistics

Schedule: 

- Presentation on data sources, types, etc. (David)
- Demos on a handful of weather data sources (Kristina)
- Address attendee questions and discuss weather data sources

## Weather data overview

See [workshop slides here](https://docs.google.com/presentation/d/e/2PACX-1vQg26WZj2vV2E_TnihD5v9pv4RG_69ssXlo4ucTfM98zHy-bNMjR9vDHn_jNAaVntZmSMTznkyCDOHa/pub?start=false&loop=true&delayms=3000). 

## Demos

### 1. Observation data from weather stations

- Find by Googling state/city/location
- Not standardized, will have to do some sleuthing
- Example: AZMet weather stations
- Referred to in presentation
- [Main informational page](https://cals.arizona.edu/azmet/az-data.htm)
- Download data manually
  - From page
  - Select Tucson Active Station
  - Click Raw Data, 2020, Daily
  - Copy data, paste into text file, and save as `azmet_tucson.csv`
- API is being developed
- Explanations of data format in "Raw Hourly and Daily Data Formats"
  - First column is year
  - Fourth column is max daily temp

```{r, message=FALSE, warning=FALSE}
azmet_tucson <- read.csv("raw_data/azmet_tucson.csv")

plot(azmet_tucson$X1, azmet_tucson$X16)
```

### 2. Interpolated gridded data

- Reanalysis data
- Lots of weather/climate sources combined
- Coverage across large areas, e.g., North America, global
- Example: Daymet
  - [Home page](https://daymet.ornl.gov/)
  - 1km x 1km grids
  - North America
  - 1950 - 2020
  - Daily values
- Download using `daymetr` R package
  - Documentation in [GitHub repo README](https://github.com/bluegreen-labs/daymetr)
  - Download for single coordinate, many coordinates, and sets of grids
- First for single location in Tucson
  - Have to do some cleaning to use
  - Column names include variable and units
  - Similar to AZMet data

```{r, message=FALSE, warning=FALSE}
library(daymetr)
library(dplyr)

daymet_tucson <- download_daymet(site = "Tucson", 
                                 lat = 32.25, 
                                 lon = -110.91, 
                                 start = 2000, 
                                 end = 2015, 
                                 internal = TRUE)

daymet_tucson_2010 <- daymet_tucson$data %>% 
  filter(year == 2010)
plot(daymet_tucson_2010$yday, daymet_tucson_2010$tmax..deg.c.)
```

- Code for getting multiple locations
  - Would need to create site csv

```{r, eval=FALSE}
download_daymet_batch(file_location = 'my_sites.csv',
                      start = 1980,
                      end = 2010,
                      internal = TRUE)
```

- Download as gridded raster files
  - NetCDF like in presentation
  - One file per year
- Read back in to plot
  - Plotting rasters can be tricky
  - Need to do more complicated stuff to plot on, e.g., Arizona map
  - Projections

```{r, results='hide', message=FALSE, warning=FALSE}
download_daymet_tiles(location = c(32.25, -110.91), 
                      start = 2010, 
                      end = 2011, 
                      param = "prcp", 
                      path = "raw_data/")
```

```{r, message=FALSE, warning=FALSE}
library(raster)
daymet_raster <- raster("raw_data/prcp_2010_11015.nc")
plot(daymet_raster)

# library(ggplot2)
# az_map <- map_data("state")
# ggplot() +
#   geom_polygon(data = az_map, aes(x = long, y = lat, group = group)) +
#   geom_raster(data = daymet_raster, aes(fill = z))
```

### 3. Forecasted climate dataset

- Different climate scenarios to choose from
- Also past forecasts, such as TerraClimate
- Example: MACA
  - [Home page](http://www.climatologylab.org/maca.html)
  - Downscale higher spatial resolution climate model predictions
  - CMIP: coupled model intercomparison project, suite of climate models
  - RCP: representative concentration pathways, have to do with radiative forcing
  - Daily data
- Use `climateR` R package to download
  - Not on CRAN
  - Get boundaries from `AOI` package
  - Turn list into RasterStack to plot
- Weird high temp values? 
  - Rasters can get really big, really fast
  - Storing values as integers instead of decimals decreases size
  - Check scale_factor in `param_meta$maca` for correction

```{r, message=FALSE, warning=FALSE}
#remotes::install_github("mikejohnson51/AOI")
#remotes::install_github("mikejohnson51/climateR")
library(AOI)
library(climateR)

maca <- getMACA(AOI = aoi_get(state = "AZ"), 
                model = "CCSM4", 
                param = "tmax", 
                scenario = "rcp45", 
                startDate = "2050-06-01", 
                endDate = "2050-06-06")

maca_raster <- stack(maca)
plot(maca_raster)
```

### 4. Climatological normals data

- Averages across time to get general assessment of climate conditions
- Useful for biological questions, e.g., characterizing habitat
- Example: bioclim
  - Part of Worldclim products
  - [main landing page](https://www.worldclim.org/data/bioclim.html)
  - Annual, seasonal, and extreme trends for biologically relevant variables
  - By year
  - Several spatial resolutions
- Download using `getData` from `raster` R package
  - These are shape files
  - Plot temps
  - Need to do scale factor correction to temp too
  - Example of extracting values for Tucson

```{r, message=FALSE, warning=FALSE}
bioclim_all <- getData(name = "worldclim", 
                       var = "bio", 
                       res = 10, 
                       path = "raw_data/")
nlayers(bioclim_all)
unlist(bioclim_all)

plot(bioclim_all$bio1)

tucson_coords <- data.frame(long = -110.91, lat = 32.25)
tucson_points <- SpatialPoints(tucson_coords)
bioclim_tucson_extract <- extract(bioclim_all, tucson_points)
bioclim_tucson <- cbind.data.frame(tucson_coords, bioclim_tucson_extract)
```
