---
title: "Making Your First R Package 📦 (Instructor Notes)"
subtitle: "ResBazAZ 2023"
author: 
  - name: Eric R. Scott
    orcid: 0000-0002-7430-7879
    affiliations: "Communications & Cyber Technologies, University of Arizona"
format: 
  html:
    toc: true
editor: visual
execute: 
  eval: false
  echo: true
---

# Introduction

[Slides](https://cct-datascience.quarto.pub/making-your-first-r-pkg-slides/#/title-slide)

::: callout-tip
Remember, even if you don't ever make an R package that goes on CRAN (or any R package at all), the skills covered in this workshop will be helpful!
:::

# Project Setup

## Create empty repo on GitHub

-   repo name = package name

-   Copy https url

## Create new R project from repo

-   New Project \> Version Control \> Git, paste in URL

## Load needed packages

```{r}
#| eval: true
library(devtools)
library(here)
```

# Git/GitHub setup

::: callout-tip
This is all a one-time per computer setup
:::

## Configure Git

Get a checklist of what to do to set things up:

```{r}
#| eval: true
git_sitrep()
```

Add a global `.gitignore` to prevent you from accidentally sharing sensitive files on GitHub:

```{r}
git_vaccinate()
```

Introduce yourself to git:

```{r}
use_git_config(
  user.name = "your name",
  user.email = "youremail@arizona.com", #email you used to sign up for GitHub account
)

```

Set a default branch name:

```{r}
#sets your default branch name to "main"
git_default_branch_configure()
```

## Git and GitHub

There are some other things to do in `git_sitrep()`, but first, let's make our first commit.
Make a commit using the git pane in RStudio.
Commit message can be something like "initial commit"

```{r}
#check how we're doing
git_sitrep()
```

Set up a personal access token (PAT).
This is necessary to securely send things between your computer and GitHub.

```{r}
create_github_token() #takes you to external website. Add description, but keep defaults
```

Store that token on your computer:

```{r}
gitcreds::gitcreds_set()
#paste in PAT from GitHub
```

Double-check that it worked:

```{r}
git_sitrep()
```

Now everything should be set (except maybe `upstream`) with no ❌s

# Package Setup

## Convert into a package

We'll use `create_package()` to convert our current project to a package.
It takes a `path` argument that we'll supply with `here::here()`

```{r}
here() #absolute path to current project
```

```{r}
create_package(here()) #converts current project to skeleton R package.  Overwrites files.
```

This will open a new RStudio window, so you can close the old one.

Re-load devtools with `library(devtools)`

## Files created

-   `.gitignore` - list files you don't want to be on GitHub

-   `.Rbuildignore` - files that you want on GitHub, but that can't be part of your R package

-   `DESCRIPTION` - contains package metadata

-   `<package name>.Rproj` - holds settings for the RStudio Project

-   `NAMESPACE` - read-only file that will be generated once we add some functions

-   `R/` - this is where you'll put R code for your package's functions

## Add a README

-   run `use_readme_md()` and edit the README.

-   You could use `use_readme_rmd()` if you want a README that executes some code, like running an example.

## Sync with GitHub

-   Commit (message = "convert to package") and push

-   Inspect GitHub repo, notice how README.md is rendered

-   Notice how files in .gitignore aren't there

# Functions

## Add a function

1.  New .R file, save as greet.R

2.  Start with super simple:

    ```{r}
    #| eval: true
    #| filename: greet.R
    greet <- function() {
      cat("Hello!")
    }
    ```

    ```{r}
    #| eval: true
    greet()
    ```

3.  `load_all()` and run `greet()`

4.  Add argument:

    ```{r}
    #| eval: true
    #| filename: greet.R
    greet <- function(name) {
      cat("Hello ", name, "!", sep = "")
    }
    ```

5.  `load_all()` and `greet("yourname")`

    ```{r}
    #| eval: true
    #| error: true
    greet("Eric")
    greet()
    ```

6.  `greet()` errors, so let's add a default

    ```{r}
    #| eval: true
    #| filename: greet.R
    greet <- function(name = "User") {
      cat("Hello ", name, "!", sep = "")
    }
    ```

7.  `load_all()` and `greet()`

    ```{r}
    #| eval: true
    greet()
    ```

## `check()`

Run `check()` and address stuff:

-   Edit author field in `DESCRIPTION`

-   run `use_mit_license()`

## Add documentation

1.  With cursor in function definition, add roxygen skeleton and fill it out
2.  `document()` then `?greet`
3.  Explore and commit changes
4.  Discuss roxygen2 fields and where to get more help on this
5.  What does `@export` mean? When would you *not* want a function to be exported?

## `check()` again

should pass

## Install

Should be able to install the package now, load it with `library()`, etc.

## Add A dependency

Let's make the output colorful with the `cli` package!

```{r}
#| eval: true
#| filename: greet.R
greet <- function(name = "User") {
  cat("Hello ", cli::col_cyan(name), "!\n", sep = "")
}
```

```{r}
#| eval: true
greet("Eric")
```

1.  Notice the syntax---`pkgname::function()`---this is *essential* so that your package knows which package functions are from!

2.  Run `check()` and notice the warning:

    ```         
    '::' or ':::' import not declared from: ‘cli’
    ```

3.  Solution: `use_package("cli")`

4.  Inspect DESCRIPTION and run `check()` again

# Data

## Add a data set

-   Read in dataset from URL (paste code into chat I guess?)

-   `use_data()` to add dataset to your package

```{r}
#read in data as an R object
baked_goods <- read.csv("https://www.ericrscott.com/files/baked_goods_long.csv")
#use_data() on the R object to add to your package
use_data(baked_goods)
```

-   Inspect output and see what is changed

    -   lazy data means the dataset is available when you load the package without having to run `data(baked_goods)`.
        Just load your package and `baked_goods` is available.

    -   .rda files are for saving R objects

-   Document the data set.
    <https://r-pkgs.org/data.html#sec-documenting-data>

    -   Create a `R/data.R`, paste in the example and edit

    -   Never `@export` a dataset, they're available through lazy loading and this isn't needed.

-   Run `document()` and check `?baked_goods`

-   Run `check()`, commit and push

# Testing

## Writing tests

-   "Unit tests" check that your code gives desired output, errors when it's supposed to, etc.

-   Set up with `use_testthat()`

-   Inspect changes

    -   `tests/`

    -   `DESCRIPTION`

-   Add a test for our function with `use_test("greet")`

-   Run test with button and with `test()` and with `check()`

-   Edit test to make sense for greet

```{r}
#| filename: test-greet.R
test_that("greet works with names", {
  expect_equal(greet("Eric"), "Hello Eric!")
})

test_that("greet works with default", {
  expect_equal(greet(), "Hello User!")
})
```

## Continuous Integration

So you've got `check()` passing on your computer, but will your package work on a different OS?
With a different version of R?

How can you find out and be alerted if changes you make will break the package for others?

Continuous Integration (CI) is a practice of automatically building your package (or running an analysis pipeline) on a variety of systems every time you make changes to your code (and push to GitHub)

We'll use GitHub Actions and `usethis` to set this up.

-   `use_github_action_check_release()` sets up a *basic* action that runs `check()` using the latest release of R running on Linux

-   `use_github_action_check_standard()` runs a more intense check of your package on linux, macOs, and Windows using the latest release of R and the development version of R.

-   Add a badge to your readme with `use_github_actions_badge()`

-   Commit, push, and inspect GitHub repo

# Sharing your Package

## Installing from GitHub

-   Add instructions to README for using `remotes::install_github("username/reponame")` to install
-   Set up a website for your package with `use_pkgdown()` and `use_pkgdown_github_pages()`

# Polishing your Package

-   Add a vignette to your package with `use_vignette()`

-   Add a code of conduct to your repo with `use_code_of_conduct()`

-   Make a checklist of what you need to do to submit your package to CRAN with `use_release_issue()`

# Appendix

## Set up build tools

> This may not be strictly necessary until you want to build packages containing C or C++ code.
> Especially if you are using RStudio, you can set this aside for now.
> The IDE will alert you and provide support once you try to do something that requires you to setup your development environment

::: callout-important
If it isn't going to work for someone, they can use Posit Cloud as a fallback.
Build tools come pre-installed there.
:::

### macOS

Check if you already have xcode command line tools installed:

``` bash
xcode-select -p
```

If you get an error instead of a path like `/Library/Developer/CommandLineTools`, then you need to install xcode command line tools with:

``` bash
xcode-select --install
```

### Windows

Install Rtools from <https://cran.r-project.org/bin/windows/Rtools/>

-   **Do *not*** select the box for "Edit the system PATH".
    devtools and RStudio should put Rtools on the `PATH` automatically when it is needed.

-   **Do** select the box for "Save version information to registry".
    It should be selected by default.
