---
title: "Playing around with the commonchemistry API"
format: html
editor: visual
---

```{r}
library(httr2)
library(tidyverse) #used later on for wrangling results
# library(listviewer) #has some functions to render lists as collapsible HTML widgets, but I want something even simpler.
```


## Search

### Build the query

```{r}
req_search <-
  request("https://commonchemistry.cas.org/api") |> 
  req_url_path_append("search") |> 
  req_url_query(q = "caffeine")
req_search
```

### Perform the request

```{r}
resp_search <- 
  req_search |> 
  req_perform()
resp_search
```

### Get the results as a list

```{r}
caffeine_search <- 
  resp_search |> resp_body_json()
caffeine_search
```

There's only one result here, so we'll simplify this list a bit by just pulling out the first (only) element of the results.

```{r}
caffeine <- caffeine_search$results[[1]]
caffeine
```

One of the things it give us is the SVG (scaleable vector graphics) code for the image, which we can display in a rendered RMarkdown or Quarto document.
Here we use a couple of "tricks" to get this to work.
We use the chunk option `output: asis` to tell Quarto to render the output as plain text and not put it inside of a box or anything.
Then we use `cat()` rather than `print()` to display the SVG code "naked" and not as a length 1 character vector.
The result is that the SVG code is interpreted as HTML and rendered as an image.

```{r}
#| output: asis
#| echo: fenced
cat(caffeine$image)
```

We could also include this image in an HTML table, like one made with `gt`.
That would be pretty cool!

## Detail

Now let's get more information with the `/detail` endpoint

### Build the query

We can pass the registry number directly from our first query into this one.

```{r}
req_detail <-
  request("https://commonchemistry.cas.org/api") |> 
  req_url_path_append("detail") |> 
  req_url_query(cas_rn = caffeine$rn)
req_detail
```

### Perform the request

```{r}
resp_detail <-
  req_detail |> 
  req_perform()
resp_detail
```

### Get the results

```{r}
caffeine_detail <- 
  resp_detail |>
  resp_body_json()
caffeine_detail
```

### Wrangle the results

Lists are hard to work with!
Let's try to turn this into a tibble.

```{r}
#| error: true
as_tibble(caffeine_detail)
```

The list isn't a nice rectangle, so it can't be easily converted into a tibble.

We can take subsets of the list where all elements are the same length and make those into data frames

Let's take all the elements with length == 1 and turn those into a data frame

```{r}
caffeine_basics <- caffeine_detail |> 
  purrr::keep(\(x) length(x) == 1) |> 
  as_tibble()
```

Ok, I gotta try that `gt` thing where we make a table with the image of the molecule:

```{r}
library(gt)
gt(caffeine_basics) |> 
  fmt_markdown(columns = c(image, molecularFormula)) #molecular formula was also HTML
```

Nice!

## Iterating

The *true* power of accessing APIs programatically is that we can iterate!
For example, we can take a list of chemicals, search for the registry number for all of them, get the results for all of them, and turn it into a single output table.

::: callout-important
Many APIs have a **rate limit** that limits how quickly you can make requests (e.g. 10 per minute).
Sometimes this rate limit is documented on the API website, sometimes you'll just get errors if you make too many requests (surprise!).
Slow down your requests with `req_throttle()`.
:::

```{r}
chemical_names <- c("caffeine", "vanillin", "bergamotene") #vanilla earl grey anyone?
```

All queries will start like this

```{r}
req_search_base <-
  request("https://commonchemistry.cas.org/api") |> 
  req_url_path_append("search")
```

Then we want to generate queries for each of the chemicals by plugging the elements of `chemical_names` into `req_url_query(q = <chemical_name>)`.
We can do that with `map()` from the `purr` package.
The first argument is the thing to iterate over—`chemical_names` in our case.
The second argument is a function to perform.
Here I've used R's "shortcut" for creating an anonymous function: `\(arg_name) do_things_to(arg_name)`

```{r}
req_list <- map(chemical_names, \(x) req_search_base |> req_url_query(q = x))
req_list
```

So now we have a list of three queries.
We can use `map()` again to perform all of them.
This time, we don't even need the anonymous function because `req_perform()` just takes the request as it's first unnamed argument.

```{r}
resp_list <- map(req_list, req_perform)
resp_list
```

And again, we can use `map` to extract all the results

```{r}
map(resp_list, resp_body_json)
```

Now here it's gotten a little complicated because we now have a list of lists.
So instead of just maping `resp_body_json()`, let's do some data extraction *inside* of `map()`.
It's usually easiest to figure out how to do this by first testing it out on just a single element.

```{r}
resp_list[[1]] |> 
  resp_body_json() |> 
  pluck("results", 1) |> #equivalent to x[[results]][[1]]
  as_tibble()
```

Then we can put this code inside an anonymous function `\(x)` and replace `resp_list[[1]]` with `x`

```{r}
df_list <- map(resp_list, \(x) {
  x |> 
    resp_body_json() |> 
    pluck("results", 1) |> #equivalent to x[[results]][[1]]
    as_tibble()
})
```

Now we have a list of tibbles that we can combine with `list_rbind()`

```{r}
search_results <- list_rbind(df_list)
search_results
```

(and then we can do the cool `gt` thing)

```{r}
gt(search_results) |> 
  fmt_markdown(columns = image)
```
