---
title: "Demystifying APIs for Researchers: Live-Coding"
subtitle: "Playing around with the commonchemistry API"
author: "Eric R. Scott"
format: 
  html:
    toc: true
editor: visual
---

These are notes for the participatory live-coding portion of a workshop.
Slides are here: <https://cct-datascience.quarto.pub/demystifying-apis-slides/>

```{r}
library(httr2)
library(tidyverse) #used later on for wrangling results
```

## Search

### Build the query

```{r}
req_search <-
  request("https://commonchemistry.cas.org/api") |> 
  req_url_path_append("search") |> 
  req_url_query(q = "caffeine")
req_search
```

### Perform the request

```{r}
resp_search <- 
  req_search |> 
  req_perform()
resp_search
```

### Get the results as a list

```{r}
caffeine_search <- 
  resp_search |> resp_body_json()
caffeine_search
```

There's only one result here, so we'll simplify this list a bit by just pulling out the first (only) element of the results.

```{r}
caffeine <- caffeine_search[["results"]][[1]]
caffeine
```

### Convert to data frame

We can convert this to a data frame or tibble.

```{r}
as_tibble(caffeine)
```

## Detail

Now let's get more information with the `/detail` endpoint

### Build the query

We can pass the registry number directly from our first query into this one.

```{r}
req_detail <-
  request("https://commonchemistry.cas.org/api") |> 
  req_url_path_append("detail") |> 
  req_url_query(cas_rn = caffeine$rn)
req_detail
```

### Perform the request

```{r}
resp_detail <-
  req_detail |> 
  req_perform()
resp_detail
```

### Get the results

```{r}
caffeine_detail <- 
  resp_detail |>
  resp_body_json()
caffeine_detail
```

### Wrangle the results

Lists are hard to work with!
Let's try to turn this into a tibble.

```{r}
#| error: true
as_tibble(caffeine_detail)
```

The list isn't a nice rectangle, so it can't be easily converted into a tibble.

We can take subsets of the list where all elements are the same length and make those into data frames

One way to "drill down" into lists is with `purrr::pluck()`

```{r}
caffeine_detail |> pluck("name")
caffeine_detail |> pluck(1)
caffeine_detail |> pluck("experimentalProperties", 1, "property")
```

We can instead use `purrr::keep()` to supply a *function* to determine which elements to extract.
For example, let's take all the elements with length == 1 and turn those into a data frame.

```{r}
caffeine_basics <- caffeine_detail |> 
  purrr::keep(\(x) length(x) == 1) |> 
  as_tibble()
```

This uses an **anonymous function**, `\(x) length(x) == 1`, which is saying "for list element `x`, is the length of `x` 1?".
`keep()` iterates like a for loop where `x` is representing `caffeine_detail[[1]]` through `caffeine_detail[[15]]`.

::: callout-tip
## Bonus

Now that we have a data frame, we can make a table with `gt` that displays the SVG data as an image of the molecule!

```{r}
library(gt)
gt(caffeine_basics) |> 
  fmt_markdown(columns = c(images, molecularFormula)) #molecular formula was also HTML
```

Nice!
This isn't a `gt` workshop, but there is a **lot** you can do with `gt` to make publication quality tables.
:::

## Iterating

The *true* power of accessing APIs programatically is that we can iterate!
For example, we can take a list of chemicals, search for the registry number for all of them, get the results for all of them, and turn it into a single output table.

::: callout-important
Many APIs have a **rate limit** that limits how quickly you can make requests (e.g. 10 per minute).
Sometimes this rate limit is documented on the API website, sometimes you'll just get errors if you make too many requests (surprise!).
Slow down your requests with `req_throttle()`.
:::

```{r}
chemical_names <- c("caffeine", "vanillin", "bergamotene") #vanilla earl grey anyone?
```

All queries will start like this

```{r}
req_search_base <-
  request("https://commonchemistry.cas.org/api") |> 
  req_url_path_append("search")
```

Then we want to generate queries for each of the chemicals by plugging the elements of `chemical_names` into `req_url_query(q = <chemical_name>)`.
We can do that with `map()` from the `purr` package.
The first argument is the thing to iterate over—`chemical_names` in our case.
The second argument is a function to perform.
Here I've again used R's "shortcut" for creating an anonymous function: `\(arg_name) do_things_to(arg_name)`

```{r}
req_list <- 
  map(chemical_names, \(x) req_search_base |> req_url_query(q = x))
req_list
```

So now we have a list of three queries.
We can use `map()` again to perform all of them.
This time, we don't even need the anonymous function because `req_perform()` just takes the request as it's first unnamed argument.

```{r}
resp_list <- map(req_list, req_perform)
resp_list
```

And again, we can use `map` to extract all the results

```{r}
map(resp_list, resp_body_json)
```

Now here it's gotten a little complicated because we now have a list of lists.
So instead of just `map()`ing `resp_body_json()`, let's do some data extraction *inside* of `map()`.
It's usually easiest to figure out how to do this by first testing it out on just a single element.

```{r}
resp_list[[1]] |> 
  resp_body_json() |> 
  pluck("results", 1) |> #equivalent to x[[results]][[1]]
  as_tibble()
```

Then we can put this code inside an anonymous function `\(x)` and replace `resp_list[[1]]` with `x`

```{r}
df_list <- map(resp_list, \(x) {
  x |> 
    resp_body_json() |> 
    pluck("results", 1) |> #equivalent to x[[results]][[1]]
    as_tibble()
})

```

Now we have a list of tibbles that we can combine with `list_rbind()`

```{r}
search_results <- list_rbind(df_list)
search_results
```

::: callout-tip
## Bonus

We can do the cool `gt` thing with multiple rows now

```{r}
gt(search_results) |> 
  fmt_markdown(columns = images)
```
:::

## Try it on your own!

As an exercise, use the three CAS registry numbers in `search_results` to search the `detail/` endpoint and make a table of chemical representations (InChI, InChIKey, SMILES, and canonical SMILES) for each molecule.
