---
title: "Data Validation in R"
author: "Eric R. Scott"
date: '2022-06-09'
output: github_document
---

## Overview

There are many opportunities for human error in data.
Let's find them before we do analysis or maybe even before we finish the field season.

## Learning Objectives

-   Understand best practices for entering data and fixing errors in data

-   Use Excel data validation tools to prevent data entry errors

-   Compare data entered by two people in R to check for data entry mistakes

-   Use the `pointblank` package to perform data validation checks

## Notes

-   **not** talking about "outliers" which are usually not appropriate to remove or change.

```{r}
library(tidyverse)
plants_raw <- read_csv("data_raw/HDP_1997_2009.txt")
plants <-
  plants_raw |> 
  #just take one plot, 2 years
  filter(year %in% 2000:2001, plot == first(plot)) |> 
  select(plot, subplot, plant_id, year, shts, ht, infl) |> 
  #pivot wider to better resemble actual data entry
  pivot_wider(names_from = year, values_from = c(shts, ht, infl), 
              names_vary = "slowest")

write_csv(plants, "data_raw/data_true.csv", na = "")

```

## Packages needed

```{r}
library(tidyverse)
library(visdat)
library(pointblank)
```

## Avoiding mistakes in data entry

### Data Validation Tools in Excel

-   Select a column (or cells) and choose Data \> Validation ...
    from menu.

-   Use "list" to restrict to specific values for categorical data

-   Use "whole number" for count data

-   Can either do before data entry or after, then use "Circle Invalid Data"

-   Watch out for dates!

    -   Even with validation turned on, Excel will auto-format anything that remotely looks like a date!

    -   Caused serious scientific errors (show paper about gene names -\> dates)

    -   Make sure no columns are set to "general" format.

    -   Disable Excel's date detection behavior

### Double-entry Method

One method is to have two people enter the same data, then compare the data programatically.
In the `data` folder, there are two versions of the example dataset—one entered by Eric and one entered by Jessica.

```{r}
eric <- read_csv("data/data_eric.csv")
jessica <- read_csv("data/data_jessica.csv")
```

We can compare them a couple of ways.
First, we can compare them visually using the `visdat` package.
This only works if the two datasets are the same dimensions

```{r}
vis_compare(eric, jessica)
```

We can also compare by finding any rows that are not identical between the two datasets with the `anti_join()` function from `dplyr`.
Adding row numbers first will make it easier to find mistakes in the csv files.

```{r}
# add rownumbers
eric    <- eric    |> mutate(row = 2:(n()+1))
jessica <- jessica |> mutate(row = 2:(n()+1))
```

```{r}
#values in `eric` that are different in `jessica`
anti_join(eric, jessica)
#values in `jessica` that are different in `eric`
anti_join(jessica, eric)
```

This finds several rows that aren't identical and have potential problems.
For example for `plant_id` 33, Eric entered "27" for the height in 2000 and Jessica entered "21".
Whoever recorded the measurement probably just had messy handwriting.
There are two rows that appear in Eric's version but not in Jessica's version.
Maybe Jessica skipped some rows or Eric accidentally entered two rows twice?
These should be double-checked on the raw data sheet.

## Catching mistakes in correctly entered data

```{r}
#after fixing data-entry errors, we get `data_resolved.csv`
plants <- read_csv("data/data_resolved.csv")
```

After correcting the data entry, there may still be mistakes in the data as a result of incorrect recording of the data (or possibly a mistake that both Jessica *and* Eric made).  
At this point, it is likely ***impossible*** to know for sure if aberrant data are truly mistakes and not just extreme values, unless you can go back to the field to check.
However, we can still flag these unexpected values to double-check the raw data or in the field next time you are there.

### The `pointblank` package

```{r}
library(pointblank)
```

[`pointblank`](https://rich-iannone.github.io/pointblank/) is an R package for data validation.  
It includes 6 (six!) data validation workflows!
Much of what it does is overkill for most of us, but the [Data Quality Reporting Workflow](https://rich-iannone.github.io/pointblank/articles/VALID-I.html) makes very cool looking tables that will help you pinpoint potential mistakes.

-   start with simple example
-   demo "CSV" button
-   then talk about action levels
-   then talk about different validation functions

```{r}
create_agent(
  tbl = plants,
  label = "plants 2000 & 2001",
) |> 
  # basic checks on column types with `col_is_*` functions
  col_is_character(columns = c(plot, subplot)) |> 
  col_is_numeric(columns = starts_with(c("shts", "ht", "infl"))) |> 
  # check values with `col_vals_*` functions
  #TODO: simplify plot or sublot to be like 5 possibilities or something easier to do programmatically
  # col_vals_in_set(columns = subplot, set = LETTERS[1:10]) |>
  col_vals_between(columns = starts_with("shts"), left = 1, right = 4, na_pass = TRUE) |> 
  col_vals_lt(columns = starts_with("ht"), 60) |> 
  # "custom" checks with col_vals_expr().  E.g. check that height is measured to nearest cm:
  col_vals_expr(~ ht_2000 %% 1 == 0) |> 
  col_vals_expr(~ ht_2001 %% 1 == 0) |> 
  interrogate()
```

(This wont show up in github doc. Prob need to use html)

-   preconditions (or just mutate and make a new agent) to check growth between years

-   publish validation to rstudio connect

-   if you want something fancy set up for your lab (automation, advanced types of validations, etc.) contact us to talk more.

## Help

Feel free to drop by the [CCT Data Science Team office hours](https://datascience.cals.arizona.edu/office-hours), which happens every Tuesday morning.
We would love to help you with your R questions about date/time, and more!

You can also [make an appointment with Eric](https://calendar.google.com/calendar/appointments/schedules/AcZssZ0xpzu3Nlo0FCIAjFFktPFzyl5Aup9-1yErSy9y8_Iq2P6DWisiYc8PhUnn6l5z74D3OtgTRPSm) to discuss this content and get troubleshooting help.
