---
title: "Dates and times in R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

### Objective

Learners will how dates and times are represented in R and be able to convert variable formats to the R standard. Once dates and times are represented correctly, learners will use the 'lubridate' package to extract components and change timezones. 
### Schedule

|Topic   |Start time |
|---|---|
|Introductions   | 11  |
|Date and time representation with POSIX   | 11:10  |
|Creating POSIXct objects from strings  | 11:15   |
|Timezones and Olson names   | 11:45  |
|Reading in dates and times from csv  | 12:00 |
|Extract time and date components with 'lubridate' and 'dplyr' | 12:15  |
|Changing timezones, two ways   | 12:30   |
|Open problem solving   | 12:45  |


## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
