---
title: "Dates and times in R"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

#### Objective

Learners will how dates and times are represented in R and be able to convert variable formats to the R standard. Once dates and times are represented correctly, learners will use the 'lubridate' package to extract components and change timezones. 

#### Schedule

|Topic   |Start time |
|---|---|
|Introductions   | 11:00 |
|1. Date representation in base R   | 11:05 |
|2. Date/time representation with POSIX | 11:25 |
|3. Timezones and Olson names   | 11:50  |
|4. Calculating time differences | 12:10  |
|5. Reading in dates and times  | 12:15 |
|6. Extract time and date components with 'lubridate' | 12:30  |
|7. Changing timezones with 'lubridate'   | 12:35   |
|Open problem solving   | 12:45  |

## Curriculum & Notes

#### 1. Date representation in base R

<i>Content based on [Berkeley Stats curriculum page](https://www.stat.berkeley.edu/~s133/dates.html)</i>

 - the base `as.Date()` function handles dates without times
 - Date objects are stored as the number of days since January 1, 1970, which can be retrieved with `as.numeric()`
 - accepts character strings and POSIXlt or POSIXct objects
 - to convert from a character string, the default format follows the ISO 8601 international standard, which is "YYYY-MM-DD"
 - if character string is <i>not</i> already in default format, the `format` argument should be specified to avoid ambiguity
  - there is an optional timzezone or `tz` argument - more on this later
 
|Type |Code   |Value |
|---|---|---|
|Day |%j | Day of year (1-366) |
|Day |%d | Day of the month (1-31) |
|Month |%m | Month (01-12) |
|Month |%b | Month(abbreviated) |
|Month |%B | Month(full name) |
|Year |%y | Year (2 digit) |
|Year |%Y | Year (4 digit) |
 
Helpful hint: If you can't recall the appropriate code, type `?strptime` into your R console and scroll down.  

Let's explore the following:
```{r, results='hide'}
x <- as.Date("2021-02-23")
x
class(x)

y <- as.Date("April 2, 2019", format = "%B %d, %Y")
y

#Try converting the following: 
"Sep 02 1995"
"19920613"
"12-Mar-12"
"2019 156"
```

#### 2. Date/time representation in POSIX

<i>Content based on [Berkeley Stats curriculum page](https://www.stat.berkeley.edu/~s133/dates.html)</i>

 - POSIX stands for '<b>P</b>ortable <b>O</b>perating <b>S</b>ystem <b>I</b>nterface', primarily for UNI<b>X</b>
 - POSIXlt stores a list of elements for second, minute, hour, day, month, and year
 - POSIXct stores date/time as number of seconds since January 1, 1970
    - POSIXct is the standard choice for dates and times in R
    - POSIX can also be used to store dates, or date/times without seconds
 - to convert character string to POSIXct, use `as.POSIXct()`
   - default ISO 8601 standard is "YYYY-MM-DD hh:mm:ss"
   - if character string is <i>not</i> already in default format, the `format` argument should be specified to avoid ambiguity
   - if timezone is not specified, `as.POSIXct()` will default to your `Sys.time()` timezone
 - to convert components of date/time into POSIXct, use `ISOdate()` or `ISOdatetime()`
   - specify numeric value of year, month, day, hour, min, sec
   - if timezone is not specified, `ISOdate()` will default to GMT
   - if timezone is not specified, `ISOdatetime()` will default to your `Sys.time()` timezone
   
 
|Type |Code   |Value |
|---|---|---|
|Hour |%H | Hour (0-24) |
|Minute |%M | Minute (0-59) |
|Second |%S | Second (0-59) |
|Dy |%j | Day of year (1-366) |
|Day |%d | Day of the month (1-31) |
|Month |%m | Month (01-12) |
|Month |%b | Month(abbreviated) |
|Month |%B | Month(full name) |
|Year |%y | Year (2 digit) |
|Year |%Y | Year (4 digit) |

Let's explore the following:
```{r, results='hide'}
x <- as.POSIXct("2021-02-23 20:30:00")
x
class(x)
str(x)
attr(x, "tzone")

y <- as.POSIXct("April 2, 2019 15:35", format = "%B %d, %Y %H:%M",
                tz = "America/Denver")
y
attr(y, "tzone")

z <- ISOdatetime(year = 2021, month = 2, day = 23, hour = 20, min = 30, sec = 0,
                 tz = "America/Phoenix")
z

# Try converting the following: 
"06:59 Sep 02 1995"
"16/Oct/2005:07:51:00"
"12-Mar-2012 0830"
2013; 8; 18
```

#### 3. Timezones and Olson names

 - both `as.Date()` and `as.POSIXct()` have an optional `tz` argument
 - there exists a [tz database](https://en.wikipedia.org/wiki/Tz_database) that compiles information about all of the world's time zones
 - the tz database name for a time zone is sometimes referred to as the Olson name
   - Check `OlsonNames()` for complete list
 - each name includes the offset from UTC (Coordinated Universal Time), including for Daylight Saving where observed
 - name is in format of Area/Location, e.g., America/Phoenix, where Area is typically the continent or ocean and Location is typically the most populous city or small island
 - for most of the United States that observes Daylight Saving, e.g. America/Los_Angeles, the Olson name automatically switches between PST and PDT
 - timezone designation affects which date is assigned when converting between POSIX and Date formats
 
Let's explore the following:
```{r, results='hide'}
x <- as.POSIXct("2021-02-23 12:30:00", tz = "America/Phoenix")
x
as.Date(x)

y <- as.POSIXct("2021-02-23 20:30:00", tz = "America/Phoenix")
y
as.Date(y)

z <- as.POSIXct("2021-02-23 20:30:00", tz = "America/Phoenix")
z_date <- as.Date(z, tz = "America/Phoenix")
z_date
```

#### 4. Calculating time differences
 - if two times (date or date/time) are subtracted, R returns the result in the form of a time difference
 - special class of 'difftime object', which includes a number and a unit
 - use the `difftime()` function specifically to obtain output in a particular unit
 - wrap in `as.numeric()` to change to obtain the number only
 
Let's explore the following:
```{r, results='hide'}
x <- as.POSIXct("2021-02-23 20:30:00", tz = "America/New_York")
y <- as.POSIXct("2021-04-23 06:30:00", tz = "America/Phoenix")
z <- as.POSIXct("2021-04-23 06:30:00", tz = "America/New_York")

y - x
str(y - x)

difftime(y, x, units = "hours")
difftime(z, x, units = "hours")
```

#### 5. Reading in dates and times

 - if using base R and `read.table` or `read.csv` to import tabular data
    - use `as.Date()` or `as.POSIXct()` to convert existing character strings
    - use `ISOdate()` or `ISOdatetime()` to convert numeric year, month, day, hour, min, and sec
    - specify `tz` argument to avoid ambiguity

Let's explore the following:
```{r, message=FALSE, warning=FALSE, results='hide'}
dat <- read.csv("data.csv")
str(dat) 

library(dplyr)
dat_clean <- dat %>%
  mutate(date_clean = as.Date(date, 
                              format = "%d-%b-%y", 
                              tz = "America/Phoenix"),
         dt_clean = as.POSIXct(dt,
                               format = "%d-%b-%y %H:%M", 
                               tz = "America/Phoenix"))
str(dat_clean)  
```

 - if using 'readr' and `read_csv()` or `read_table()` to import tabular data
    - readr parses each column of strings into a vector 
    - under the hood, strings in standard formats are automatically parsed into date, time, or date/time column types 
    - otherwise, the format must be specified in the `col_types` argument
    - timezone can be specified in the `locale` argument to avoid ambiguity, or else your `Sys.time()` timezone will be the default

Let's explore the following:
```{r, message=FALSE, warning=FALSE, results='hide'}
library(readr)
dat <- read_csv("data.csv")

dat <- read_csv("data.csv", 
                col_types = list(date = col_date(format = "%d-%b-%y"),
                                 dt = col_datetime(format = "%d-%b-%y %H:%M")),
                locale = locale(tz = "America/Phoenix"))

str(dat)
attr(dat$dt, "tzone")
``` 

#### 6. Extract time and date components with 'lubridate'

<i>Content based on [R for Data Science](https://r4ds.had.co.nz/dates-and-times.html#date-time-components) section and date and times</i>

 - once data are read in correctly in POSIX format, components can be extracted with functions from the `lubridate()` package
 - `year()`, `month()`, `hour()`, `minute()`, and `second()` are self-explanatory
 - day can be extracted as:
   - day of the month with `mday()`
   - day of the year with `yday()`
   - day of the week with `wday()`
   
Let's explore the following: 
```{r, results='hide'}
library(lubridate)

dat %>%
  mutate(year = year(dt),
         doy = yday(dt),
         hour = hour(dt),
         min = minute(dt))

```   
   
#### 7. Changing timezones with 'lubridate'  

<i>Content based on [R for Data Science](https://r4ds.had.co.nz/dates-and-times.html#date-time-components) section and date and times</i>

 - there are two distinct intentions with changing time zones:
  1. Keep the numeric value (e.g., seconds since January 1, 1970) the same, but alter how it is displayed
  2. Change the underlying numeric value because the original timezone was incorrect
 - 'lubridate' has clear functions for each instance
 - specify the `tzone` argument in `with_tz()` to change the time zone displayed
 - specify the `tzone` argument in `force_tz()` to correct the time zone displayed

Let's explore the following: 
```{r, results='hide'}
x <- as.POSIXct("2021-02-23 20:30:00", tz = "America/New_York") 

x1 <- with_tz(x, tzone = "America/Phoenix")
x; x1
difftime(x, x1)

x2 <- force_tz(x, tzone = "America/Phoenix")
x; x2
difftime(x, x2)
```

## Help

Feel free to drop by the [CCT Data Science Team office
hours](https://datascience.cals.arizona.edu/office-hours), which happen
every Tuesday morning. We would love to help you with your R questions about date/time, and more!

You can also [make an appointment with
Jessica](https://calendar.google.com/calendar/u/0/selfsched?sstoken=UU5paG1Xb1hlUXR5fGRlZmF1bHR8NjM0M2VmZmRkZWYzYTM2N2I5ZGM4M2UyYzEyN2U4MDA)
to discuss this content and get troubleshooting help.