---
title: "Exploring the Wide World of ggplot2 Extensions"
subtitle: "Instructor notes"
authors:
  - Eric R. Scott
  - Kristina Riemer
format: 
  pdf: default
  html: default
editor: visual
execute: 
  eval: false
  echo: true
---

```{r}
library(palmerpenguins)
library(ggplot2)
```

# Intro

Go through introductory [slides](https://cct-datascience.quarto.pub/exploring-the-wide-world-of-ggplot2-extensions/) to start.
Spend time looking through the ggplot extension gallery and ask if there are any particular packages that look useful to folks.
Get people to share their discoveries.

# Demonstrations

Go through vignette or other short demo of several extension packages.

## `gganimate`

```{r}
library(gganimate)
```

-   Animate plots and save as gifs

-   Takes some getting used to lingo

-   Can control: what changes over time with `transition_*`, how the view of the axes changes with `view_*`, how much "memory" of the change there is, with `shadow_*`, and how data appears and disappears with `enter_*` and `exit_*`

-   Use `?` with functions to figure out what label name is

-   These penguin plots don't really make sense because the data isn't same individuals over the years; should set group in `transition_states`

-   Mention `animate()` for more control

-   Resources:

    -   [Bat data slides](https://goodekat.github.io/presentations/2019-isugg-gganimate-spooky/slides.html#1)

    -   [Penguins code source](https://adiradaniel.netlify.app/post/gganimate/)

## `esquisse`

```{r}
library(esquisse)
library(dplyr)
```

-   "Esquisse" means rough first sketch in French

-   `esquisse` [vignette](https://dreamrs.github.io/esquisse/articles/get-started.html)

-   Point and click to generate ggplot plots from data

-   Good for exploring datasets (look at data too)

-   Still have to modify dataset outside of tool, e.g., recoding factors

```{r}
penguins <- penguins %>% na.omit()
esquisser(penguins)
```

-   Helpful things: color codes variables, suggests plot types

-   Save plot as image file, or copy and paste code into console (uses `tidyverse`)

-   Can add to Shiny app to let non-R users create ggplots of data

## `plotly`

```{r}
library(plotly)
```

-   `plotly` R package is interface to a JavaScript library

-   Only important to know that because if you search for `plotly` help, sometimes you'll get JavaScript code examples.

-   `plotly` *kind of* uses the grammar of graphics, but building a plot from scratch is tricky

-   `plotly` functions mostly take lists are arguments, which makes it very difficult to figure out what default values are or even what arguments functions can take.

-   `ggplotly()` **transforms** `ggplot` objects into interactive `plotly` plots.
    90% of the time it gets you 90% of the way there

```{r}
p <- 
  ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", color = "Penguin Species") +
  theme_bw()

ggplotly(p)
```

By default, you get:

-   Drag a box to zoom, double-click to zoom out

-   Tooltips on hover

-   Ability to hide and show points by clicking on the legend

-   Toolbar with tools including ability to download a PNG of the plot

This works in R Markdown or Quarto documents (i.e. there's not a way to make these plots "stand-alone"---they're rendered as HTML)

### Types of plots

Works for most (all?) built-in geoms.
Works for some plots made with ggplot2 extensions also.
Demo a few different geoms like `geom_boxplot()`, `geom_hisotgram()`, `geom_smooth()`

### Customize tooltip

You can customize the info displayed in the tooltip with `ggplotly` .
Give it additional aesthetic `text` to include something *only* as a tooltip.

```{r}
p <- 
  ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  #Ignore unknown aesthetics warning
  geom_point(aes(text = sex)) +
  labs(x = "Bill length (mm)", y = "Bill depth (mm)", color = "Penguin Species") +
  theme_bw()

ggplotly(p, tooltip = c("text", "bill_depth_mm", "bill_length_mm"))
```

### Animations

You can use `ggplotly()` to make interactive animated plots by using the `frame` aesthetic

```{r}
library(gapminder) #dataset
head(gapminder)
p2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  scale_x_log10()

ggplotly(p2)

```

## `ggrepel`

The `ggrepel` package is helpful for directly labeling plots, especially when labels would otherwise overlap.

```{r}
#make rownames into columns
dat <- mtcars |> tibble::rownames_to_column(var = "car")
dat
p3 <- 
  ggplot(dat,
  aes(wt, mpg, label = car, colour = factor(cyl))) +
  geom_point()
#without ggrepel
p3 + geom_text()
p3 + geom_label()
library(ggrepel)
p3 + geom_text_repel()
p3 + geom_label_repel()
```

### Adjust appearance of labels

By default, little segments are drawn connecting labels to points only when labels are far enough away.
This can be adjusted.

```{r}
p3 + geom_label_repel(min.segment.length = 0)
```

### Overlapping labels

You see a warning about labels being removed due to overlaps.
Make text smaller, plot larger, adjust `max.overlaps` , force labels further apart, make it try harder to find positions that don't overlap.

```{r}
#don't usually need to do all of these things, but they're some options
p4 <- 
  p3 + 
  geom_label_repel(
    min.segment.length = 0,
    #smaller labels
    size = 3, 
    label.padding = 0.15,
    #draw labels even if overlapping with things
    max.overlaps = 15, #default is 10, increase to get more labels drawn
    #force labels further from point
    force = 5,
    #ask it to try harder
    max.time = 2,
    max.iter = 15000,
    show.legend = FALSE
  )
p4
```

### Labels are random

Labels are positioned randomly, so they're different every time a plot is rendered

```{r}
#| layout-ncol: 2
p4
p4
```

You can `set.seed()` at the top of your script to ensure reproducibility.

```{r}
#| layout-ncol: 2
set.seed(123)
p4
set.seed(123)
p4
```

### Label just some points

You can label just select points

```{r}
cars <- c("Volvo 142E", "Merc 230")
ggplot(dat, aes(wt, mpg, label = ifelse(car %in% cars, car, ""))) +
  geom_point(color = "red") +
  geom_label_repel(min.segment.length = 0, nudge_x = 0.5, nudge_y = 3)
```

# Wrap-up

Go through "where to find help" slides.
Plug next workshop (our fall reproducibility series)

Ask if there are any other remaining `ggplot2`-related roadblocks people are experiencing and workshop it live.
