---
title: "Exploring the Wide World of ggplot2 Extensions"
author: 
  - "Eric R. Scott"
  - "Kristina Riemer"
format: 
  revealjs:
    slide-number: true
    chalkboard: true
    link-external-icon: true
    link-external-newwindow: true
    theme: 
      - "../_extensions/cct-datascience/uaz/theme.scss"
      - "custom.scss"
    logo: logo.png
execute: 
  echo: true
editor: visual
---

## Learning Objectives

-   Understand where to find packages that extend `ggplot2`
-   Make animated plots with `gganimate`
-   Interactively create plots with `esquisse`
-   Convert `ggplot2` plots to interactive `plotly` plots
-   Label data with `ggrepel`

## Packages

```{r}
library(ggplot2)
library(palmerpenguins)
```

## Extensions to `ggplot2`

There are many kinds of extensions to `ggplot2`, but many of them can be put into one of two "bins"

1.  Packages that add new layers
2.  Packages that wrap multiple layers for convenience

## Packages that extend `ggplot2`

Some packages add new layers or themes that can be combined with standard `ggplot2` functions.

```{r}
#| code-line-numbers: "|1,4"
library(ggbeeswarm)
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  geom_beeswarm()

```

::: notes
`geom_beeswarm()` can easily be added to other `geom_`s
:::

## Packages that wrap `ggplot2`

Other packages provide convenience "all-in-one" functions that combine multiple layers for you.

```{r}
#| code-line-numbers: "|1-3"
library(ggpubr)
ggstripchart(penguins, x = "species", y = "body_mass_g",
             add = "boxplot") +
  scale_y_continuous("Body Mass (g)", n.breaks = 10)
```

::: notes
Notice that `ggstripchart()` totally replaces `ggplot(aes(…))` .
However, it still makes a ggplot object that you can add layers to.
:::

## Finding Extensions

-   Browse the `ggplot2` [extensions gallery](https://exts.ggplot2.tidyverse.org/gallery/)

-   Check out the [Awesome `ggplot2`](https://github.com/erikgahner/awesome-ggplot2#readme) list

-   [The R Graph Gallery](https://r-graph-gallery.com/)

-   Google search with "ggplot2" keyword

# Package Demos

# Anything we missed?

Got a `ggplot2` question we didn't cover in this workshop series?
Let's figure it out together!

## Getting Help

-   Website / documentation for extension packages

-   Our [drop-in hours](https://datascience.cct.arizona.edu/drop-in-hours)

-   [UA Data Science Slack](https://join.slack.com/t/uadatascience/shared_invite/zt-1xjxht9k0-E489WA6axO_SRVeKQh3cXg)
